Miniproject - Combinatorics

### 2021-12-02 Combinatorics/submission.tex
Started part 5 of combinatorics

\section{Part IV}
\begin{example}
    Let $\Pi_m = \{ \pi \subseteq \powerset(n) \ | \ \text{$\pi$ partition} \}$ be the poset of partitions ordered by refinement.
    Then $\Pi$ is a graded poset with rank function
    \begin{equation*}
        r: \Pi \to \{ 0, ..., n - 1 \}, \quad \pi \mapsto n - |\pi|
    \end{equation*}
\end{example}
\begin{proof}
    First of all, have $1 \leq |X| \leq n$ for all subsets $X \subseteq \powerset(n)$, hence the function $r$ is well-defined.
    Next we show that for all $x < y$ have $r(x) < r(y)$.
    However, if $x$ is a proper refinement of $y$, then clearly $|x| > |y|$, so $r(x) < r(y)$.

    Now assume there is a maximal chain $\pi_1 < ... < \pi_t$ in $\Pi$.
    As the chain is maximal and $\Pi$ has the maximal element $[n]$, we may assume that $\pi_t = [n]$ and thus $r(\pi_t) = n - 1$.
    Similarly, we may assume that $p_1 = \powerset(n)$ and so $r(\pi_1) = 0$.
    We have to show that $t = n$, and so it suffices to show that $r(\pi_i) = 1 + r(\pi_{i - 1})$ for all $0 < i \leq t$.

    Assume not, i.e. $|\pi_{i - 1}| \geq 2 + |\pi_i|$ for some $0 < i \leq t$.
    Since $\pi_{i - 1}$ is a refinement of $\pi_i$, we find that
    \begin{equation*}
        \pi_{i - 1} = \{ A_{11}, ..., A_{1m_1}, \ ..., \ A_{n1}, ..., A_{nm_n} \} \quad \text{with} \quad \pi_i = \Bigl\{ \bigcup_{j \leq m_1} A_{1j}, ..., \bigcup_{j \leq m_n} A_{nj} \Bigr\}
    \end{equation*}
    where $n = |\pi_i|$ and $\sum_j m_j = |\pi_{i - 1}| \geq 2 + n$.
    Hence, for at least two different $s, r \leq n$ we have $m_s, m_r \geq 2$.
    Now define
    \begin{equation*}
        \tilde{\pi} := \{ A_{jk} \ | \ j \neq s \} \cup \Bigl\{ \bigcup_{j \leq m_s} A_{sj} \Bigr\}
    \end{equation*}
    Then clearly $\pi_{i - 1} < \tilde{\pi} \leq \pi_i$.
    However, we know that $\bigcup_j A_{rj} \in \pi_i$ but $\bigcup_j A_{rj} \notin \tilde{\pi}$.
    Thus $\pi_{i - 1} < \tilde{\pi} < \pi_i$ and we found a longer chain
    \begin{equation*}
        \pi_1 < ... < \pi_{i - 1} < \tilde{\pi} < \pi_i < ... < \pi_t
    \end{equation*}
    which contradicts the assumed maximality. The claim follows.
\end{proof}
\begin{proposition}
    \label{prop:partitions_recursion}
    For the number $S(m, k) := |\{ \pi \in \Pi \ | \ |\pi| = k\}|$ we have recursion
    \begin{equation*}
        S(m, 1) = 1 \quad \text{and} \quad S(m, k) = \sum_{n = 1}^{m - 1} {m - 1 \choose n} S(n, k - 1)
    \end{equation*}
    where $1 \leq k \leq m$.
\end{proposition}
\begin{proof}
    Use induction on $k$. 
    The base case $k = 1$ is trivial, so let $k > 1$.
    For any partition $\pi \in \Pi_m$, denote by $A(1, \pi)$ the unique set $A \in \pi$ with $1 \in A$.
    We find
    \begin{align*}
        S(m, k) =& |\{ \pi \in \Pi_m \ | \ |\pi| = k \}| = \Bigl| \bigcup_{\emptyset \neq A \subseteq [m], \ 1 \in A} \{ \pi \in \Pi_m \ | \ |\pi| = k, A(1, \pi) = A \} \Bigr| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_m \ | \ |\pi| = k, A(1, \pi) = A \}| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_m \ | \ |\pi \setminus \{A\}| = k - 1, A(1, \pi) = A\}| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_{m - |A|} \ | \ |\pi| = k - 1 \}| \\
        =& \sum_{i = 1}^m \ \sum_{A \subseteq [m], \ |A| = i, \ 1 \in A} |\{ \pi \in \Pi_{m - i} \ | \ |\pi| = k - 1 \} \\
        =& \sum_{i = 1}^m {m - 1 \choose i - 1} S(m - i, k - 1) = \sum_{i = 0}^{m - 1} {m - 1 \choose m - i - 1} S(i, k - 1) \\
        =& \sum_{i = 1}^{m - 1} {m - 1 \choose i} S(i, k - 1)
    \end{align*}
    as for $k > 1$ have that $S(0, k) = 0$.
\end{proof}
\begin{example}
    Have for $m \geq 3$ that
    \begin{equation*}
        S(m, 2) = 2^{m - 1} - 1 \quad \text{and} \quad S(m, 3) = \frac 1 2 (3^{m - 1} - 2^m + 1)
    \end{equation*}

@@

\end{document}

>>>>

\section{Part IV}
\begin{example}
    Let $\Pi_m = \{ \pi \subseteq \powerset(n) \ | \ \text{$\pi$ partition} \}$ be the poset of partitions ordered by refinement.
    Then $\Pi_m$ is a graded poset with rank function
    \begin{equation*}
        r: \Pi_m \to \{ 0, ..., n - 1 \}, \quad \pi \mapsto n - |\pi|
    \end{equation*}
\end{example}
\begin{proof}
    First of all, have $1 \leq |X| \leq n$ for all subsets $X \subseteq \powerset(n)$, hence the function $r$ is well-defined.
    Next we show that for all $x < y$ have $r(x) < r(y)$.
    However, if $x$ is a proper refinement of $y$, then clearly $|x| > |y|$, so $r(x) < r(y)$.

    Now assume there is a maximal chain $\pi_1 < ... < \pi_t$ in $\Pi_m$.
    As the chain is maximal and $\Pi_m$ has the largest element $[n]$, we may assume that $\pi_t = [n]$ and thus $r(\pi_t) = n - 1$.
    Similarly, we may assume that $p_1 = \powerset(n)$ and so $r(\pi_1) = 0$.
    We have to show that $t = n$, and so it suffices to show that $r(\pi_i) = 1 + r(\pi_{i - 1})$ for all $0 < i \leq t$.

    Assume not, i.e. $|\pi_{i - 1}| \geq 2 + |\pi_i|$ for some $0 < i \leq t$.
    Since $\pi_{i - 1}$ is a refinement of $\pi_i$, we find that
    \begin{equation*}
        \pi_{i - 1} = \{ A_{11}, ..., A_{1m_1}, \ ..., \ A_{n1}, ..., A_{nm_n} \} \quad \text{with} \quad \pi_i = \Bigl\{ \bigcup_{j \leq m_1} A_{1j}, ..., \bigcup_{j \leq m_n} A_{nj} \Bigr\}
    \end{equation*}
    where $n = |\pi_i|$ and $\sum_j m_j = |\pi_{i - 1}| \geq 2 + n$.
    Hence, there is one $s \leq n$ with $m_s \geq 3$ or two different $s, r \leq n$ with $m_s, m_r \geq 2$.

    In the first case, consider
    \begin{equation*}
        \tilde{\pi} := \{ A_{jk} \ | \ j \neq s \} \cup \Bigl\{ A_{s1}, \bigcup_{2 \leq j \leq m_s} A_{sj} \Bigr\}
    \end{equation*}
    and in the second case, consider
    \begin{equation*}
        \tilde{\pi} := \{ A_{jk} \ | \ j \neq s \} \cup \Bigl\{ \bigcup_{j \leq m_s} A_{sj} \Bigr\}
    \end{equation*}
    Then $\pi_{i - 1} < \tilde{\pi} < \pi_i$ and we found a longer chain
    \begin{equation*}
        \pi_1 < ... < \pi_{i - 1} < \tilde{\pi} < \pi_i < ... < \pi_t
    \end{equation*}
    which contradicts the assumed maximality. The claim follows.
\end{proof}
\begin{proposition}
    \label{prop:partitions_recursion}
    For the number $S(m, k) := |\{ \pi \in \Pi_m \ | \ |\pi| = k\}|$ we have recursion
    \begin{equation*}
        S(m, 1) = 1 \quad \text{and} \quad S(m, k) = \sum_{n = 1}^{m - 1} {m - 1 \choose n} S(n, k - 1)
    \end{equation*}
    where $1 \leq k \leq m$.
\end{proposition}
\begin{proof}
    Use induction on $k$. 
    The base case $k = 1$ is trivial, so let $k > 1$.
    For any partition $\pi \in \Pi_m$, denote by $A(1, \pi)$ the unique set $A \in \pi$ with $1 \in A$.
    We find
    \begin{align*}
        S(m, k) =& |\{ \pi \in \Pi_m \ | \ |\pi| = k \}| = \Bigl| \bigcup_{\emptyset \neq A \subseteq [m], \ 1 \in A} \{ \pi \in \Pi_m \ | \ |\pi| = k, A(1, \pi) = A \} \Bigr| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_m \ | \ |\pi| = k, A(1, \pi) = A \}| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_m \ | \ |\pi \setminus \{A\}| = k - 1, A(1, \pi) = A\}| \\
        =& \sum_{\emptyset \neq A \subseteq [m], \ 1 \in A} |\{ \pi \in \Pi_{m - |A|} \ | \ |\pi| = k - 1 \}| \\
        =& \sum_{i = 1}^m \ \sum_{A \subseteq [m], \ |A| = i, \ 1 \in A} |\{ \pi \in \Pi_{m - i} \ | \ |\pi| = k - 1 \} \\
        =& \sum_{i = 1}^m {m - 1 \choose i - 1} S(m - i, k - 1) = \sum_{i = 0}^{m - 1} {m - 1 \choose m - i - 1} S(i, k - 1) \\
        =& \sum_{i = 1}^{m - 1} {m - 1 \choose i} S(i, k - 1)
    \end{align*}
    as for $k > 1$ have that $S(0, k) = 0$.
\end{proof}
\begin{example}
    Have for $m \geq 3$ that
    \begin{equation*}
        S(m, 2) = 2^{m - 1} - 1 \quad \text{and} \quad S(m, 3) = \frac 1 2 (3^{m - 1} - 2^m + 1)
    \end{equation*}

@@

\section{Part V}
\begin{example}
    Let $\mathcal{P}_{k, d} := \{ 0, ..., k \}^d$ partially ordered by elementwise ordering.
    Then $\mathcal{P}_{k, d}$ is a graded poset with rank function
    \begin{equation*}
        r: \mathcal{P}_{k, d} \to \{ 0, ..., kd \}, \quad a \mapsto \sum_i a_i
    \end{equation*}
\end{example}
\begin{proof}
    First, we show that for $a < b$ have $r(a) < r(b)$.
    If $a < b$, then we find that $a_j \leq b_j$ for all $j$ and $a_i \neq b_i$, so $a_i < b_i$ for some $i$.
    Hence
    \begin{equation*}
        r(a) = \sum_j a_j = a_i + \sum_{j \neq i} a_j \leq a_i + \sum_{j \neq i} b_j < b_i + \sum_{j \neq i} b_j = \sum_j b_j = r(b)
    \end{equation*}
    Now assume that we have a maximal chain $a_1 < ... < a_t$ in $\mathcal{P}_{k, d}$.
    Since $\mathcal{P}_{k, d}$ has the smallest element $0 = (0, ..., 0)$ and the largest element $k = (k, ..., k)$, we see that $p_1 = 0$ and $p_t = k$.
    We want to show $t = kd + 1$, so it suffices to show that $r(a_i) = r(a_{i - 1}) + 1$ for all $0 < i \leq t$.

    Assume not, then 
    \begin{equation*}
        r(a_i) - r(a_{i - 1}) = \sum_j a_{ij} - a_{(i - 1)j} \geq 2
    \end{equation*}
    For all $j$ we have $a_{ij} \geq a_{(i - 1)j}$ and so there is $s \leq d$ with $a_{is} - a_{(i - 1)s} \geq 2$ or there are different $r, s \leq d$ with $a_{ir} - a_{(i - 1)r}, a_{is} - a_{(i - 1)s} \geq 1$.
    In both cases, find
    \begin{equation*}
        \tilde{a} \in \mathcal{P}_{k, d} \ \text{defined by} \ \tilde{a}_j = \begin{cases}
            a_{(i - 1)j} & \text{if $j \neq s$} \\
            a_{(i - 1)j} + 1 & \text{otherwise}
        \end{cases}
    \end{equation*}
    with $a_{i - 1} < \tilde{a} < a_i$. However, this gives a longer chain
    \begin{equation*}
        a_1 < ... < a_{i - 1} < \tilde{a} < a_i < ... < a_t
    \end{equation*}
    which contradicts the assumed maximality. The claim follows.
\end{proof}
\end{document}
